### 2017-07-26 applications/clawpack/advection/2d/swirl_cons/notes/fwave.tex
(clawpack/swirl (cons)) UPdate write-up

\begin{document}
\section{Introduction}
\section{Numerical approaches}

\subsection{Scalar approach to conservation}
In \cite{le:2002}, LeVeque describes a solution to the conservation law
\begin{align}
q_t + (uq)_x &= 0 \\
\end{align}
where $u = u(x)$ is a spatially varying velocity field, and $q = q(x,t)$ is a conserved  tracer in  the flow field.  Using the analogy of boxes moving along a series of conveyor belts operating
at speeds prescribed by $u(x)$, LeVeque describes a solution involving two waves. One moves at local speeds given by each conveyor belts, and stationary waves at the interfaces between conveyor belts.  To see this, we let suppose we have two conveyor belts. The left moves at speed $u_\ell$ and the right moves at speed $u_r$.  Each conveyor belt has a density of boxes, given by $q_\ell$ and $\q_r$. 
If we have $u_\ell > u_r > 0$, boxes will pile up on the right conveyor belt  as faster moving boxes from the left are transfered to the slower moving conveyor belt on the right.  If $0 < u_\ell < u_r$, the density of boxes on the right conveyor belt will drop as faster moving boxes on the right move away from boxes arriving at a slower rate from the left.  In either case, we expect a jump 
in box density both at the wave moving at speed $u_r$ along the right conveyor belt, and at the interface between the left and right conveyor belts.

Mathematically, we can see that single wave is not enough by considering the Riemann problem at an interface between conveyor belts.   To solve the Riemann problem, we might try to connect left and 
right states $q_\ell$ and $q_r$ by a single wave moving at some speed $s$ satisfying
\begin{equation}
u_r q_r - u_\ell q_\ell = s(q_r - q_\ell)
\end{equation}
Since $u_r \ne u_\ell$ in general, a single wave that connects the left and 
right states $q_\ell$ and $q_r$ does not exist.  Instead, we write the conservation 
law as
\begin{align*}
q_t + u_r q_x & = 0, \qquad \mbox{in the left cell} \\
q_t + u_\ell q_x & = 0, \qquad \mbox{in the right cell}
\end{align*}
and introduce two waves, one moving at speed zero, and the second moving at speed $u_r > 0$.  
We seek an intermediate state $q^*$ that can be used to connect to left and right states.
Across the wave moving at speed $u_r > 0$, we impose the condition
\begin{equation}
u_r q_r - u^* q^* = u_r(q_r - q^*)
\end{equation}
to get $u^* = u_r$.  Across the stationary wave, we impose a flux continuity condition
\begin{equation}
u^* q^* - u_\ell q_\ell = 0
\end{equation}
to get
\begin{equation*}
q^* = \frac{u_\ell q_\ell}{u_r}.
\end{equation*}
The waves moving at speeds 0 and speed $u_r > 0$ are given by
\begin{align*}
\mathcal{W}_1 & \equiv q^* - q_\ell = \frac{u_\ell q_\ell}{u_r} - q_\ell = 
-\frac{u_r - u_\ell}{u_r}q_\ell\\
\mathcal{W}_2 & \equiv q_r - q^* = q_r - \frac{u_\ell q_\ell}{u_r} = 
\frac{1}{u_r}\left(u_r q_r - u_\ell q_\ell\right)
\end{align*}
The fluctuations are then defined in terms of these waves as
\begin{align*}
\mathcal{A}^+ \Delta Q & \equiv u_r \mathcal{W}_1 = u_r q_r - u_\ell q_\ell \\
\mathcal{A}^- \Delta Q & \equiv 0
\end{align*}
The sum of the fluctuations satisfy the conservation requirement
\begin{equation*}
\mathcal{A}^+ \Delta Q + \mathcal{A}^- \Delta Q = f(q_r) - f(q_\ell)
\end{equation*}

\subsection{2-wave decomposition approach.}
Consider the following conservation law for the conservative form of the scalar advection equation.
\begin{align}
q_t + (uq)_x &= 0 \\
u_t + (au)_x & = 0
\end{align}
for conserved quantity $q(x,t)$, and  $u = u(x,t)$ is a prescribed velocity field.  We will only 
interested in the case $a = 0$ so that $u = u(x)$ depends on space only.

@@

Following the above approach, we can also find waves for the case $\widehat{u} < a = 0$.   Combining these two cases, we get the following Clawpack Riemann solver.

@@

\begin{minipage}{\textwidth}
\begin{verbatim}
# Riemann solver for wave decomposition approach
def rpn2(ql,qr,ul,ur):
    uavg = (ul + ur)/2        
    uhat = sign(uavg)*max([abs(uavg),tol])
    qhat = (ql + qr)/2
    wave = qr - ql + (ur-ul)*qhat/uhat
    if uavg == 0:
        amdq = -ul*qhat     # amdq+apdq = (ur-ul)*qhat
        apdq = ur*qhat
        s = uavg
    else:
        apdq = max([uhat,0])*wave
        amdq = min([uhat,0])*wave        
        s = uhat
    return wave,s,amdq,apdq
\end{verbatim}
\end{minipage}

>>>>

\newcommand{\Fig}[1]{Figure \ref{fig:#1}}

\begin{document}
\section{Introduction}
\section{Numerical approaches}
The variable coefficient conservative form of the transport equation arises in many areas of
geophysical flow modeling.  For example, the transport of pollutants, volcanic ash or 
other conserved tracer component must often transported using prescibed reanalysis 
wind field data.

In \cite{le:2002}, LeVeque describes several approaches to solving the conservative, 
variable coefficient advection equation, given by
\begin{align}
q_t + (uq)_x = 0 \\
\end{align}
In his book, LeVeque describes several approaches to solving this equation, although
there is a question about what to do in the case where the velocity field changes sign. 
In this case, there is a loss of hyperbolicity that cannot be easily addressed by the 
wave propagation algorithms implemented in ClawPack.  Here, we propose reasonable approaches
to handling this case.

In the following, we provide details for four of the approaches described, and pay
particular attention to ways to treat the change of sign that can lead to loss of 
hyperbolicity.  The accuracy obtained from such methods is second order for velocity
fields with no sign changes.  For velocity fields with sign changes, the accuracy can be
significantly degraded, depending on the use of limiters, and whether the velocity field 
converges or diverges at the sign change.

\subsection{Scalar approach to conservation}
In \cite{le:2002}, LeVeque describes a solution to the conservation law
\begin{align}
q_t + (uq)_x = 0 \\
\end{align}
where $u = u(x)$ is a spatially varying velocity field, and $q = q(x,t)$ is a conserved  tracer in  the flow field.  Using the analogy of boxes moving along a series of conveyor belts operating
at speeds prescribed by $u(x)$, LeVeque describes a solution involving two waves. One moves at local speeds given by each conveyor belts, and stationary waves at the interfaces between conveyor belts.  To see this, we let suppose we have two conveyor belts. The left moves at speed $u_\ell$ and the right moves at speed $u_r$.  Each conveyor belt has a density of boxes, given by $q_\ell$ and $q_r$. 
If we have $u_\ell > u_r > 0$, boxes will pile up on the right conveyor belt  as faster moving boxes from the left are transfered to the slower moving conveyor belt on the right.  If $0 < u_\ell < u_r$, the density of boxes on the right conveyor belt will drop as faster moving boxes on the right move away from boxes arriving at a slower rate from the left.  In either case, we expect a jump 
in box density both at the wave moving at speed $u_r$ along the right conveyor belt, and at the interface between the left and right conveyor belts.

We can see that single wave is not enough by considering the Riemann problem at an interface between conveyor belts.   To solve the Riemann problem, we might try to connect left and 
right states $q_\ell$ and $q_r$ by a single wave moving at some speed $s$ satisfying
\begin{equation}
u_r q_r - u_\ell q_\ell = s(q_r - q_\ell)
\end{equation}
Since $u_r \ne u_\ell$ in general, a single wave that connects the left and 
right states $q_\ell$ and $q_r$ does not exist.  Instead, we write the conservation 
law as
\begin{align*}
q_t + u_r q_x & = 0, \qquad \mbox{in the left cell} \\
q_t + u_\ell q_x & = 0, \qquad \mbox{in the right cell}
\end{align*}
and introduce two waves, one moving at speed zero, and the second moving at speed $u_r > 0$.  
We seek an intermediate state $q^*$ that can be used to connect to left and right states.
Across the wave moving at speed $u_r > 0$, we impose the condition
\begin{equation}
u_r q_r - u^* q^* = u_r(q_r - q^*)
\end{equation}
to get $u^* = u_r$.  Across the stationary wave, we impose a flux continuity condition
\begin{equation}
u^* q^* - u_\ell q_\ell = 0
\end{equation}
to get
\begin{equation*}
q^* = \frac{u_\ell q_\ell}{u_r}.
\end{equation*}
The waves moving at speeds 0 and speed $u_r > 0$ are given by
\begin{align*}
\mathcal{W}_1 & \equiv q^* - q_\ell = \frac{u_\ell q_\ell}{u_r} - q_\ell = 
-\frac{u_r - u_\ell}{u_r}q_\ell\\
\mathcal{W}_2 & \equiv q_r - q^* = q_r - \frac{u_\ell q_\ell}{u_r} = 
\frac{1}{u_r}\left(u_r q_r - u_\ell q_\ell\right)
\end{align*}
The fluctuations are then defined in terms of these waves as
\begin{align*}
\mathcal{A}^+ \Delta Q & \equiv u_r \mathcal{W}_1 = u_r q_r - u_\ell q_\ell \\
\mathcal{A}^- \Delta Q & \equiv 0
\end{align*}
The sum of the fluctuations satisfy the conservation requirement
\begin{equation*}
\mathcal{A}^+ \Delta Q + \mathcal{A}^- \Delta Q = f(q_r) - f(q_\ell)
\end{equation*}

The Riemann solver for this approach is described in \Fig{qstar}.

\begin{figure}
\begin{center}
\begin{minipage}{0.5\textwidth}
\vspace{0.5cm}
\begin{verbatim}
# Riemann solver for "conveyor belt" approach
def rpn2(ql,qr,ul,ur):
    if ur > 0 and ul > 0:
        wave = (ur*qr - ul*ql)/ur
        amdq = 0
        apdq = ur*qr - ul*ql
        s = ur
    else if ur < 0 and ul < 0:
        wave = (ur*qr - ul*ql)/ul
        amdq = ur*qr - ul*ql
        apdq = 0
        s = ul
    else:
        uavg = (ur+ul)/2.0
        uhat = max(abs(uvg),1e-12)
        wave = (ur*qr - ul*ql)/uhat
        amdq = -ul*ql
        apdq = ur*qr
        s = uhat

    return wave,s,amdq,apdq
\end{verbatim}
\end{minipage}
\end{center}
\caption{Algorithm for cell-centered velocities based on scalar equations.}
\label{fig:qstar}
\end{figure}

\subsection{2-wave decomposition approach.}
Problem 13.11 in \cite{le:2002}
Consider the following conservation law for the conservative form of the scalar advection equation.
\begin{align}
q_t + (uq)_x &= 0 \\
u_t + (au)_x & = 0
\end{align}
for conserved quantity $q(x,t)$, and  $u = u(x,t)$ is a prescribed velocity field.  We will only 
interested in the case $a = 0$ so that $u = u(x)$ depends on space only.

@@

The Riemann solver for this approach is described in \Fig{wave_decomp}.

\begin{figure}
\begin{center}
\begin{minipage}{0.5\textwidth}
\vspace{0.25cm}
\begin{verbatim}
# Riemann solver for wave decomposition approach
def rpn2(ql,qr,ul,ur):
    qhat = (ql + qr)/2.0

    uavg = (ul + ur)/2.0      
    if uavg == 0:
        uhat = 1e-12
    else
        uhat = uavg

    s = uavg
    wave = (ur*qr - ul*ql)/uhat

    if uavg == 0:
        amdq = -ul*qhat     # amdq+apdq = (ur-ul)*qhat
        apdq = ur*qhat
    else:
        apdq = max([uhat,0])*wave
        amdq = min([uhat,0])*wave

    return wave,s,amdq,apdq
\end{verbatim}
\end{minipage}
\end{center}
\caption{2-wave decomposition approach to handling conservative advection equation.}
\label{fig:wave_decomp}
\end{figure}

@@

\subsection{Edge velocities}
In some sense, the simplest approach is to assume that the velocity field at cell edges is available. 
In this case, we can easily evaluate a flux at cell interfaces, and formulate the problem 
in classic flux form, where the flux is defined as
\begin{equation*}
F_{i-1/2} = \left\{\begin{array}{cc} 
u_\ell q & u_\ell, u_r < 0 \\
u_r q & u_r, u_\ell > 0
\end{array}
\right.
\end{equation*}
where
